---
title: "Importing and Cleaning Data from a Large Dataset"
author: "Davena Mgbeokwere"
date: "2025-02-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
 
#Load Libraries 

```{r load_libraries}

library (tidyverse) # for importing and cleaning data
library (summarytools) #contains dfsummary() function which provides summary of stats for all variables selected

```

#Import and Combine Datasets

```{r import_data}

# show_col_types = FALSE to silence message about columns. default shows first few column names and guidance on how to retrieve column specification

raw_demographics <- 
  read_csv("C:/Users/School/OneDrive - Georgia State University/Documents/December 2025/Research/MA thesis/thesis moderation analysis/thesis moderation analysis/data/RAW_thesis_demographics.csv", show_col_types = FALSE)

raw_covariates <- 
  read_csv("C:/Users/School/OneDrive - Georgia State University/Documents/December 2025/Research/MA thesis/thesis moderation analysis/thesis moderation analysis/data/RAW_thesis_covariates.csv", show_col_types = FALSE)

```


```{r create_sample_dataset}

sample_raw_demo <- raw_demographics %>% 
  sample_frac (.25) %>% #randomly sample 25 of the entire dataset
  mutate(
    project = case_when( #case_when () used when creating single variable with multiple conditions
      M2ID > 0 ~"M2",
      MRID > 0 ~"Refresher"),
    .keep = c("unused"),
    .after = 1
    ) %>% 
  select (-SAMPLMAJ)
  

sample_raw_covariates <- 
  sample_raw_demo %>% 
  select(MIDUSID) %>% 
  left_join(raw_covariates, by="MIDUSID") %>%
  select (-SAMPLMAJ:-MRID)


```

##Dataset 1

```{r rename_variables_demographic_dataset}

# For some variables, there were 4 variables that needed to be combined into one variable. To create a new variable combining these four variables, I used case_when () which allows you to have multiple conditions. 
# For variables where I only needed to combine two variables, I used if_else () to create a single variable.

demo_data <- sample_raw_demo %>% 
  mutate( # function to create a new variable
    MIDUSID = as.character (MIDUSID),
    sex = case_when(B1PRSEX > 0 ~ B1PRSEX, # M2P1 sex
                    RA1PRSEX > 0 ~ RA1PRSEX, # MR 1 sex
                    RAACRSEX > 0 ~ RAACRSEX, # Milwaukee refresher sex
                    BACRSEX > 0 ~ BACRSEX), # Milwaukee 1 sex
    race1 = case_when(B1PF7A > 0 ~ B1PF7A, # M2P1 first race choice
                      BACF7A > 0 ~ BACF7A, # Milwaukee 1 first race choice
                      RA1PF7A > 0 ~ RA1PF7A, # MR 1 first race choice
                      RAACF7A > 0 ~ RAACF7A), # Milwaukee refresher first race choice
    race2 = case_when(B1PF7B > 0 ~ B1PF7B, # M2P1 second race choice
                      BACF7B > 0 ~ BACF7B, # Milwaukee 1 second race choice
                      RA1PF7B > 0 ~ RA1PF7B, # MR 1 second race choice
                      RAACF7B >0 ~ RAACF7B), # Milwaukee R second race choice
    int_age = case_when(B1PAGE_M2 > 0 ~ B1PAGE_M2,# M2P1 interview age
                        RA1PRAGE > 0 ~ RA1PRAGE, # MR 1 interview age
                        RAACRAGE > 0 ~ RAACRAGE, # Milwaukee R interview age
                        BACRAGE > 0 ~ BACRAGE), # Milwaukee 1 interview age
    bio_age = ifelse(is.na(B4ZAGE), RA4ZAGE, B4ZAGE), # age at biomarker visit
    biomarker = case_when(bio_age > 0 ~ 1, is.na(bio_age) ~ 0), # biomarker participant? 0 = No, 1 = yes
    educ = case_when(B1PB1 > 0 ~ B1PB1, # M2P1 education
                     BACB1 > 0 ~ BACB1, # Milwaukee 1 education
                     RA1PB1 > 0 ~ RA1PB1, # MR 1 education
                     RAACB1 >0 ~ RAACB1), # Milwaukee R education
    stress = ifelse (is.na (B4QPS_PS), RA4QPS_PS,B4QPS_PS), #PSS score
    logRMSSD1 = ifelse(is.na(B4VB1LRM),RA4VB1LRM, B4VB1LRM), # baseline 1 RMSSD
    logRMSSD2 = ifelse(is.na(B4VB2LRM),RA4VB2LRM, B4VB2LRM), # baseline 2 RMSSD
    ) %>% 
  select (MIDUSID:logRMSSD2)

```

##Dataset 2

```{r rename_variables_covariates_dataset}

# For this dataset, I used to same functions to combine variables. 

cov_data <- sample_raw_covariates %>%
  mutate(
    MIDUSID = as.character (MIDUSID),
    sleepdur = if_else(is.na(B4SSQ_S3), RA4SSQ_S3, B4SSQ_S3),
    sleepqual=if_else (is.na(B4SSQ_S1),RA4SSQ_S1, B4SSQ_S1),
    BMI = if_else(is.na (B4PBMI), RA4PBMI, B4PBMI),
    avgsystolic = if_else (is.na(B4P1GS), RA4P1GS, B4P1GS),
    avgdiastolic = if_else (is.na(B4P1GD), RA4P1GD, B4P1GD), 
    MASQgendisdep = if_else (is.na(B4QMA_D),RA4QMA_D, B4QMA_D),
    MASQanxarousal = if_else(is.na(B4QMA_AA),RA4QMA_AA, B4QMA_AA),
    MASQposaffect = if_else(is.na(B4QMA_PA),RA4QMA_PA, B4QMA_PA),
    MASQgendisanx = if_else (is.na(B4QMA_A), RA4QMA_A, B4QMA_A),
    MASQlossint = if_else (is.na(B4QMA_LI), RA4QMA_LI, B4QMA_LI),
    CESDscore = if_else(is.na (B4QCESD), RA4QCESD, B4QCESD), 
    spielangerexpin = if_else (is.na(B4QAE_AI),RA4QAE_AI, B4QAE_AI),
    spielangerexpout = if_else (is.na (B4QAE_AO), RA4QAE_AO, B4QAE_AO), 
    spielangerexpcon = if_else (is.na (B4QAE_AC), RA4QAE_AC, B4QAE_AC),
    spieltraitanxinv = if_else (is.na (B4QTA_AX), RA4QTA_AX, B4QTA_AX),
    soc_anxiety = if_else (is.na (B4QSA_SA), RA4QSA_SA, B4QSA_SA),
    evertobacco = if_else (is.na (B4O7), RA4H37, B4O7),
    currtobacco = if_else(is.na (B4H29), RA4H43A, B4H29),
    eversmoke = if_else (is.na (B4H26), RA4H38, B4H26),
    currsmoke = ifelse (is.na (B4H26A), RA4H38A, B4H26A),
    drinkpastmonth = if_else (is.na (B4H33), RA4H49, B4H33),
    daily5drinks = if_else (is.na (B4H37), RA4H53, B4H37),
    weeklyMETminutes= if_else (is.na (B4HMETMW), RA4HMETMW,B4HMETMW),
    heartdisease = if_else (is.na (B4H1A), RA4H1A, B4H1A),
    highBP = if_else (is.na (B4H1B), RA4H1B, B4H1B),
    heartmurmur = if_else (is.na (B4H1E), RA4H1E, B4H1E),
    tiastroke = if_else (is.na (B4H1F), RA4H1F, B4H1F),
    cholesterol = if_else (is.na (B4H1H), RA4H1H, B4H1H),
    diabetes = if_else (is.na (B4H1I), RA4H1I, B4H1I),
    alcoholism = if_else (is.na (B4H1U), RA4H1U, B4H1U),
    depression = if_else (is.na (B4H1V), RA4H1V, B4H1V),
    waisthip = if_else (is.na (B4PWHR), RA4PWHR, B4PWHR),
    resprate = if_else (is.na (B4P1E), RA4P1E, B4P1E),
    .keep = "unused",
    .after = 1) %>% 
# I wanted to create a single variable from 15 variables that met a single condition. First, I renamed then variables using rename_with (). I wanted to rename the medication variables to start with "med" followed by a number. To do this, I used paste0 () to create the name for the variable. seq_along (.) numbers the variables.
  rename_with (., ~ paste0("med", seq_along(.)), starts_with('B4XPICD9M')) %>%
  select (MIDUSID:resprate, starts_with("med"))


```

##Combine Datasets 

```{r combine_datasets}

merged_data <- 
  left_join(demo_data, cov_data, by = "MIDUSID") 

```

# Clean Data

```{r explore_combined_data}

dfSummary (merged_data) #use to obtain summary data for each variable

```

After reviewing the summary data, the first thing I decided to address was re-coding the values for missing data to NA so the values will not be included in calculations. I wasn't sure of the best way to do this and probably made it more complicated than I needed to but several values were used to indicate missing values based on the range of the variable.

```{r Change_Missing_to_NA}

# There may have been a better way to do this but I needed to remove all missing values across all variables without removing valid values so I created to separate values groups and variable groups

# create objects for values ---------------------------------------------------------------

missing_values_high <- 
  c(97, 98,99)

missing_values_low <- 
  c(7,8,9)

# create objects for variables -----------------------------------------------------------

missing_variables_high <- c( "educ", "resprate", "stress", 'logRMSSD1','logRMSSD2', "daily5drinks")

missing_variables_low <- 
  c("heartdisease", "highBP", "heartmurmur", "tiastroke", "cholesterol", 
                "diabetes", "alcoholism", "depression", "sleepdur", "sleepqual",
                "race1", "race2", "currsmoke", "currtobacco",  "eversmoke" , 
                "evertobacco", "drinkpastmonth", 'logRMSSD1','logRMSSD2',
    "soc_anxiety", "waisthip")

# replace missing values with NA---------------------------------------------------------

missing_converted <- 
  merged_data %>% 
  mutate(
    across( # applies mutation to multiple variables/columns
      c(missing_variables_high, MASQgendisdep : spieltraitanxinv), # can use this or all_of ()
      ~replace (., . %in% missing_values_high, NA)), # replace in dataset-variables in object - with NA
    across(
      all_of(missing_variables_low), #can use this or c ()
      ~replace(., . %in% missing_values_low, NA)) # replace in dataset-variables in object - with NA
    )


```


```{r explore_combined_data}

dfSummary (missing_converted)

```

After converting all missing values to NA, I needed to create new variables, change some numeric variables to factors, and re-code some factor variables
Need to create smoking variable 
need to create single drinking variable 
Need to create single race variable
need to create BP meds variable
need to compute RMSSD variable 

```{r Create_New_Variables}

updated_merged <- missing_converted %>%
  
# create blood pressure variable 
  mutate (bloodpres = avgsystolic/avgdiastolic) %>% 
  
# create tobacco user variable from multiple variables. 
  
# I started with the easiest logic formulas, which were coding the current user and non-user. 
# Then, I spent some time figuring out how to code the quitters. After that, there were still  cases
# that weren't categorized. After some trial-and-error, I figure out how to set up the code to categorize them correctly.
  
  mutate (tobacco_user = case_when(evertobacco == 2 & eversmoke == 2 ~ "nonuser", # NO coded as 2. 
                                  currtobacco == 1 | currsmoke == 1 ~ "user",  # YES coded as 1.
                                  eversmoke == 1 & currsmoke == 2 ~ "quit",
                                  evertobacco == 1 & currtobacco == 2 ~ "quit",
                                  is.na (evertobacco) & eversmoke== 2 & 
                                    currtobacco == 2 ~ "nonuser",
                                  evertobacco == 2 & is.na (currtobacco) & 
                                    is.na(eversmoke) & is.na (currsmoke) ~"nonuser"), 
          .after = 30) %>%
  
# create drinking variable
  mutate (occ_5_drinks = if_else (drinkpastmonth == 3 | drinkpastmonth == 2, 0, daily5drinks),
                                .after = 32) %>%
# create new race variable 
  mutate (proj_race = case_when (race1 == 1 & is.na (race2) ~ race1, race1 == 2 & is.na (race2)~
                               race1, .default = NA), .after = 4)

# create blood pressure medication variable ------------------------------------

meds_function <- function (x) x == 401

updated_merged_bp <- 
  updated_merged %>% 
# I looked up how to create a single blood pressure medication variable. The most efficient way to create a variable that applies the same condition to multiple variables that I found was to use apply (). Reads select all variables that start "med", 1 = apply over all select rows, for any that meet the condition x== 401
   mutate(bp_meds = apply(select(., starts_with("med")), 1, function(x) any(meds_function(x))), 
          bp_meds_2 = case_when(bp_meds == 1 ~ 1, .default = 0)) 
  select (-med1:-med15)
          
#create average of two columns for HRV------------------------------------------
  
avghrv_updated_merged <- updated_merged_bp %>% 
  rowwise() %>% #calculate average of two variables across columns. This took awhile with the full dataset but not with the sample dataset
  mutate(avglogRMSSD = mean(c_across(c('logRMSSD1', 'logRMSSD2')),na.rm=TRUE), 
         .after = 12)

```

Need to convert some variables to factors
 
```{r Convert_Numeric_to_Factor}

#factors to convert to variables 

factorvariables <- c ("project", "alcoholism", "biomarker", "bp_meds", "cholesterol", "currsmoke", "currtobacco", "depression","diabetes","drinkpastmonth", "educ" , "eversmoke" , "evertobacco","heartdisease", "sleepqual", "sleepdur" ,"heartmurmur", "highBP" , "race1", "race2" , "proj_race" ,"sex", "tiastroke", "tobaccouser")

factor_updated <- avghrv_updated_merged %>%
                        mutate_at (factorvariables, as.factor)


```


```{r Recode_Factor_Variables}

educlevels <- c (1:12)

clean_full_data <- factor_updated %>%  

mutate (full_race = case_when (is.na (race1) ~ NA,
                               race1 == 1 & is.na (race2) ~ "W",
                               race1 == 2 & is.na (race2)~ "B",
                           .default="M"), .after = 4) %>% 
mutate (rev_race = factor (dplyr::recode (proj_race, "1" = "1", "2" = "0"), 
                            levels = c(0, 1), labels = c ("Black", "White")),
        .after = 4) %>% 
mutate (rev_sex = factor (dplyr::recode (sex, "1" = "1", "2" = "0"), 
                            levels = c(0, 1), labels = c ("Female", "Male")),
        .after = 2) %>% 
mutate (educ2 = factor (case_when(educ %in% c("1", "2", "3") ~ "0",
                                  educ %in% c("4", "5") ~ "1",
                                  educ %in% c("6", "7") ~ "2",
                                  educ == "8" ~ '3',
                                  educ == "9" ~ "4",
                                  educ == "10" ~ "5",
                                  educ %in% c("11", "12") ~ "6"), 
                        levels = c (0:6), 
                        labels = c('< GED/HS diploma', 'GED/HS diploma', 
                                   'some college', '2 yr degree', 
                                   '4 yr degree' , 'some grad school', 
                                   'graduate degree'), ordered = T),
        .after = 9) %>% 
mutate (stress2 = case_when (stress <=13 ~ "low", stress > 13 & stress <= 26 ~ "moderate", 
                             stress > 27 ~ "high"),
        .after = 10) %>% 
mutate (rev_sleepdur = factor(dplyr::recode(sleepdur,
                                   '0' = '3', '1' = '2','2'= '1','3' = '0'),  
                        levels = c(0, 1, 2, 3),
                        labels = 
                      c("< 5 HOURS", "5-6 HOURS", "6-7 HOURS", "> 7 HOURS"),
                      ordered = T),
        rev_sleepqual = factor(dplyr::recode(sleepqual,
                                             '0' ='3', '1' = '2','2' = '1', '3' = '0'),  
                               levels = c(0, 1, 2, 3), 
                               labels = c("< Very Bad", "Fairly Bad", "Fairly Good", 
                                          "Very Good"), ordered = T), .after = 15) %>% 
mutate(across(c(depression, alcoholism, tiastroke, heartmurmur, heartdisease),
              ~ dplyr::recode(., "1" = "1", "2" = "0","3" = "0"
)))
  

```


```{r Create_Project_Dataset }

proj_data <- clean_full_data %>% 
                filter (biomarker==1, !is.na (proj_race)) %>% 
                dplyr::select (-biomarker)

              
```
